Prediction of Early Periprosthetic Joint Infection after Total Hip Arthroplasty: Development and External Validation of a Multivariable Model Based on Observational Registry data from Sweden and Denmark

### 2021-10-15 reports/article.Rmd
Submission to Osteoarthritis & Cartilage. Fix #61.

---
title: "Prediction of Early Periprosthetic Joint Infection after Total Hip Arthroplasty: Development and External Validation of a Multivariable Model Based on Observational Registry data from Sweden and Denmark"
author: Erik Bülow,^1,2^ Ute Hahn,^3,4^ Ina Trolle Andersen,^3^ Ola Rolfson,^1,2^  Alma Besic Pedersen,^3,5^ Nils P. Hailer^6^
date: '`r Sys.Date()`'
bibliography: 
#  H:/zotero_lib.bib
  C:\\Users\\erik_\\Documents\\library.bib
csl: bmj.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotm
    toc: false
    df_print: kable
    number_sections: false
---

@@

**Word count:** XXX

@@

**DESIGN:** Cohort study.

**SETTING:** Register data from Sweden for model derivation and (from Denmark for external) validation.

**PARTICIPANTS:** `r nmbr(N_shar)` patients from the Swedish Hip Arthroplasty Register, and `r nmbr(N_dk)` patients from the Danish Hip Arthroplasty Register operated with elective THA.

**MAIN OUTCOME MEASURE:** Logistic LASSO regression with bootstrap ranking was used to develop a risk prediction model for PJI within 90 days based on the Swedish cohort, followed by external validation on the Danish cohort.

@@

**CONCLUSION:** A new PJI prediction model based on easily accessible data available before THA, has been developed and externally validated. The model had superior discriminatory ability compared to ASA class alone or more complex comorbidity indices and had good calibration. We provide a web-based calculator (https://erikbulow.shinyapps.io/thamortpred/) to facilitate shared decision making by patients and surgeons.

# Summary box

## What is already known on this topic

- Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA).
- Attempts have been made to develop individualized risk prediction models for clinical usage in the US and Australia but none of those models has gained general acceptance due to limited samples, low discriminating ability, sub-optimal calibration or absence of external model validation.

## What this study adds
- Our prediction model was derived on a Swedish cohort and externally validated on a Danish cohort
- Our model is based on easily accessible data that are available before THA
- Compared to more complex comorbidity indices or to ASA class, our model had superior discriminatory ability and good calibration
- We provide a simple web calculator for easy use in clinical practice (https://erikbulow.shinyapps.io/thamortpred/).

@@

Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA), mainly due to its association with increased mortality, re-operation risk, long-term antibiotic treatment, and often persistently impaired quality of life.[@Zimmerli2006; @Kapadia2016; @Shohat2019] Many risk factors for the development of PJI are identified, with anaemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male sex are examples of non-modifiable risk factors.[@Ridgeway2005; @Lai2007; @Malinzak2009; @Ong2009; @Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015] Individualized PJI risk stratification would be a highly relevant tool in the pre-operative assessment of patients scheduled for THA. Both the American Society for Anesthesiologists' (ASA) classification or considerably more complex comorbidity indices can be used for risk profiling, but their use is rather challenging. Although high ASA class is a risk factor for worse clinical outcomes, it was never developed for the purpose of predicting PJI [@Haynes1995; @Berbari2012]. The Charlson[@Charlson1987] and Elixhauser[@Elixhauser1998] comorbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions according to the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA. The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified along the Anatomical Therapeutic Chemical (ATC) Classification.[@Sloan2003; @Pratt2018] Several attempts have been made at developing prognostic individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first.[@Edelstein2015; @Wingert2016a] Additional models have since been derived on US or Australian cohorts, but none has gained general acceptance.[@Wingert2016a; @Bozic2013a; @Inacio2015a] This is mainly due to often limited, sometimes single-center based samples, low discriminating ability or sub-optimal calibration of the prediction models, or the absence of external model validation. We therefore aimed to develop a prognostic prediction model of PJI within 90 days after THA that is based on easily accessible data that are available in the setting of clinical decision-making. Furthermore, we aimed to validate the model both internally in Sweden, as well as externally on a Danish national cohort.

@@

The derivation cohort was obtained from SHAR `r period`, with the choice of 2008 as the starting year because registration of ASA class and body mass index (BMI) was initiated then. Deterministic data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration.[@Ludvigsson2009; @Cnudde2016] The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of THA fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI prior to surgery, ASA class at the time of surgery, type of hospital, and year of surgery, were obtained from the SHAR. Data on comorbidities recognized during one year prior to index surgery, were obtained from the National Patient Register (NPR) where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to ICD-10 together with admission and discharge dates. Data on marital status and the educational level of patients were obtained from the longitudinal integration database for health insurance and labour market studies (LISA) from Statistics Sweden. The SHAR has completeness of 96-98 % and 100 % coverage (82 hospitals in 2020),[@Karrholm2018] the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %.[@Ludvigsson2011] Death dates were linked from the national population register. Only the last operated hip was considered for patients with bilateral THA.[@Bulow2020] Patients with missing information on ASA class, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a measurement above 50. Patients with missing information on educational levels or the type of hospital performing index surgery were also excluded, as were children under the age of 18.

@@

All surgery related variables, ASA and BMI were obtained from the DHR. The DHR is a clinical quality database established in 1995 on all primary and revision THA surgeries, with registration completeness of more than 95 % during the last decade for both primary and revision THAs.[@Gundtoft2016] All hospitals performing THA in Denmark (40 in 2019) are included in the register. 
The Danish National Patient Register holds data on ICD-10 codes as well as all procedures codes for hospitalizations and outpatient visits at Danish public and private hospitals.[@Schmidt2015] Data on education and civil status were obtained from Statistics Denmark.[@zotero-2587]

@@

The Charlson and Elixhauser comorbidity indices are based on the presence of ICD-10 codes grouped into `r nrow(coder::charlson)` and `r nrow(coder::elixhauser)` categories each.[@Quan2005] Such codes were identified from reported in- and outpatient hospital visits recorded in the national patient registers during one year prior to surgery. Diagnostic categories from those indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Tab. \@ref(tab:tabcategorization)).

@@

PJI was defined by the occurrence of relevant ICD-10 or procedural Nordic Medico-Procedural Committee (NOMESCO)-codes recorded in the national patient registers for hospital visits within 90 days of THA (Tab. \@ref(tab:infcodes)), or if the patient was reported to SHAR or DHR as re-operated due to infection within the same time frame.

@@

Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO).[@Guo2015; @Baranowski2020; @Garland2021] Age was normalized prior to modelling, and the estimated coefficients thus indicated variable importance on the same scale as categorical variables. BMI was divided into four categories based on the six intervals proposed by the World Health Organization (WHO; BMI $<25$: under- or normal weight, $25-29.9$: overweight, $30-34.9$: class I obesity, and BMI $\geq 35$: class II-III obesity).

Ten-fold cross validation was performed with a range of potential penalty values ($\lambda_i$:s) in the LASSO model, keeping the $\lambda$ that minimized the mean cross-validated deviance. 100 bootstrap samples were then drawn from the observed data set.[@Austin2004]

@@

Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison, as well as multivariable models combining those measures with age, sex and BMI (10 models in total). The Rx Risk V index is a comorbidity index based on medical prescriptions coded by the Anatomical Therapeutic Chemical (ATC) classification system during one year prior to surgery. This data was retrieved from the Swedish and Danish[@Kildemoes2011] national prescription registers and classified according to Pratt et al.[@Pratt2018]

@@

The main, reduced and the 10 reference models were used to predict the probability of PJI within 90 days for patients derived from the SHAR (internal validation). Receiver operating characteristic (ROC) curves and their corresponding areas under the curve (AUC) were used to describe discrimination with 95 % confidence intervals based on an asymptotic approximation suggested by DeLong et al.[@DeLong1988] The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed. Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration.[@Nattino2016] Coefficients of determination were estimated by Nagelkerke's $R^2$ for each model. The baseline model with the lowest $R^2$ ($R^2_0$) was compared to the other $k = 1, \ldots, 11$ models, by $1 - R^2_0 / R^2_k$. This is the fraction of new information, or the proportion of added explainable variation for model $k$ compared to baseline.[@Harrell2018]

@@

We used `r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages coder,[@Bulow2020b] decoder, furrr, pROC, rms, tidymodels and tidyverse. We build a shiny-based web calculator to aid clinical usage (https://erikbulow.shinyapps.io/thamortpred/).

@@

Results from the 100 bootstrap replicates for variable selection were mostly coherent. Diagnosis, BMI and some comorbidities where selected each time. Age was selected 82 out of 100 times, but some factors, such as civil status and type of hospital, were only sporadically selected (Tab. \@ref(tab:brlprop)).

@@

Both Sweden and Denmark have strived for mass higher education.[@Isopahkala-Bouret2018] This is, to a higher degree achieved via the university sector in Sweden, while the educational system is more diverse in Denmark.[@Kim2002] National comparisons of educational levels are therefore difficult, since the Danish student register does not include adult- or continuing education.[@zotero-2598] Patient distribution among ASA classes were similar in both countries, but the comorbidity burden seems to differ. ICD-10 and NOMESCO codes are recorded for both in- and outpatient hospital visits in the national patient registries in both countries, but the recording of (optional) secondary diagnoses is 23 % in Denmark,[@Schmidt2015] compared to 47 % in the Swedish cohort. Serious comorbidities such as cancer, are therefore reported equally in both countries, while for example arterial hypertension, is less documented in Denmark. Also, patients with chronic conditions, such as diabetes, are more commonly treated in primary care settings in Denmark, and are therefore less reported to their national patient register.[@Schmidt2015] The RxRisk V index is a weighted index, which is difficult to compare without further analysis. Prescription patterns might differ by indications based on different national guidelines, different subsidization, and different medications sold over-the-counter. Also, the Danish, but not the Swedish, national prescription register, includes prescriptions from nursery homes and other long-term care facilities. Taken together, the similarities between the countries might be more pronounced than indicated by Tab. \@ref(tab:tab1).

However, the difference in the mode of THA fixation, with a dominance of cemented components in Sweden and much more widespread usage of uncemented components in Denmark is tangible and has been described elsewhere.[@Bunyoz2020]

@@

Cancer seems to confer an increased risk of PJI in several observational studies,[@Bozic2012; @Everhart2013; @Poultsides2013; @Wingert2016a] but this comorbidity was not included in our model. It nevertheless appeared in `r subset(coefs_selected_tab, variable == "Cancer", "n", drop = TRUE)` of the 100 bootstrap replicates (Tab. \@ref(tab:brlprop)), and could have been included in a less parsimonious model. Neither diabetes was included in our model, but has been in previous risk calculators.[@Wingert2016a; @Tan2018; @Inacio2015] On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery.[@Namba2012]

Information on peri-operative hyperglycaemia, morning blood glucose, HBA1c,[@Mraovic2011; @Iorio2012] or low preoperative haemoglobin,[@Rasouli2014a] could have improved our model. But although we had access to the ICD codes defining the presence of anaemia, we had no information on actual laboratory findings. Use of subcutaneously applied Tumour Necrosis Factor-alpha blockers,[@Gilson2010; @Momohara2011] intra-articular steroid application[@Cancienne2015] or operating time,[@Maoz2015; @Willis-Owen2010; @Berbari2012] are also associated with the risk of PJI after THA. However, we chose to only consider factors available before surgery and are thus able to use the model as a tool for shared decision making using the implemented web calculator. This is a strength of our model.

We also confirmed that, although mentioned as risk factors for surgical site infections after joint arthroplasty,[@Rasouli2014a; @Inacio2015a] prediction models based on the comorbidity indices developed by Charlson or Elixhauser and the RxRisk V index, resulted in poor predictive power. This agrees with previous findings describing that these measures of comorbidity rather poorly predict mortality and patient reported outcomes after THA.[@Bulow2017; @Bulow2019] However, when used in conjunction with information on age and sex, the AUC for these comorbidity indices increased, although never quite reaching the AUC of our model (Fig. \@ref(fig:aucci) and Tab. \@ref(tab:fni)).

@@

The selected variables are nevertheless coherent with previous studies. Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons.[@Pedersen2010a; @Dale2012; @Bozic2012; @Inacio2015a] Obesity is associated with an increased risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with an increased risk of developing surgical site infections or PJI after total hip or knee arthroplasty.[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013; @Tan2018] CNS disease (cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease) has also been previously associated with PJI,[@Bozic2012] as have male sex,[@Dale2012; @Wingert2016a; @Tan2018] liver disease [@Deleuran2015] and ASA class.[@Berbari2012; @Pulido2008; @Namba2012]

THA fixation using cement without antibiotics confers an increased risk of PJI,[@Dale2012] but such cement brands were not in clinical use in Sweden during the studied period. We found no indication that the type of fixation was associated with the risk of PJI, and the mode of THA fixation was also not selected as a potential predictor variable in any of the iterations during model development.

@@

# Contribution of authors

@@

# Data sharing

Access to the model derivation cohort is granted after permission from the Västra Götaland Region, as the central personal data controller agency for the SHAR, in accordance with the European General Data Protection Regulation (GDPR), as well as Swedish laws and regulations (https://registercentrum.se/vara-tjaenster/datauttag/p/B1eeV66QE).

# Disclosure

@@

@@

>>>>

---
title: "Prediction of Early Periprosthetic Joint Infection after Total Hip Arthroplasty: Development and External Validation of a Multivariable Model Based on Observational Registry data from Sweden and Denmark"
author: Erik Bülow,^1,2^ Ute Hahn,^3,4^ Ina Trolle Andersen,^3^ Ola Rolfson,^1,2^  Alma Besic Pedersen,^3,5^ Nils P. Hailer^6^
date: '`r Sys.Date()`'
bibliography: 
#  H:/zotero_lib.bib
  C:\\Users\\erik_\\Documents\\library.bib
csl: oac.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotm
    toc: false
    df_print: kable
    number_sections: false
---

@@

**Word count:** XXX **(MAX 4000!!!)**

@@

**DESIGN:** Cohort study with register data. `r nmbr(N_shar)` patients from the Swedish Hip Arthroplasty Register used for model derivation and internal validation, and `r nmbr(N_dk)` patients from the Danish Hip Arthroplasty Register used for external validation. Logistic LASSO regression with bootstrap ranking was used for model development.

@@

**CONCLUSIONS:** A new PJI prediction model based on easily accessible data available before THA, has been developed and externally validated. The model had superior discriminatory ability compared to ASA class alone or more complex comorbidity indices and had good calibration. We provide a web-based calculator (https://erikbulow.shinyapps.io/thamortpred/) to facilitate shared decision making by patients and surgeons.

# Key words

Total hip arthroplasty, prosthesis joint infection, prediction, model development, model validation

# Running title

Prediction model for PJI after THA

@@

Periprosthetic joint infection (PJI) is the most devastating of early complications after total hip arthroplasty (THA), mainly due to its association with increased mortality, re-operation risk, long-term antibiotic treatment, and often persistently impaired quality of life[@Kapadia2016]. Many risk factors for the development of PJI are identified, with anaemia, diabetes, and obesity being important, to some extent modifiable risk factors, whereas advanced age and male sex are examples of non-modifiable risk factors[@Pedersen2010; @Bozic2012; @Jamsen2012; @Rasouli2014a; @Maoz2015]. Individualized PJI risk stratification would be a highly relevant tool in the pre-operative assessment of patients scheduled for THA. Both the American Society for Anesthesiologists' (ASA) classification or considerably more complex comorbidity indices can be used for risk profiling, but their use is rather challenging. Although high ASA class is a risk factor for worse clinical outcomes, it was never developed for the purpose of predicting PJI[ @Berbari2012]. The Charlson and Elixhauser comorbidity indices are based on `r nrow(coder::charlson)` or `r nrow(coder::elixhauser)` conditions according to the International Classification of Diseases (ICD), respectively, but they are cumbersome to use in a clinical setting, and they were also not designed for prediction of specific adverse events after THA. The same is true for the Rx Risk V-classification with `r nrow(coder::rxriskv)` conditions codified along the Anatomical Therapeutic Chemical (ATC) Classification[@Pratt2018]. Several attempts have been made at developing prognostic individualized risk prediction models for clinical usage, with the Universal American College of Surgeons National Surgical Quality Improvement Program (ACS NSQIP) Surgical Risk Calculator being one of the first[@Edelstein2015; @Wingert2016a]. Additional models have since been derived on US or Australian cohorts, but none has gained general acceptance[@Wingert2016a; @Bozic2013a; @Inacio2015a]. This is mainly due to often limited, sometimes single-center based samples, low discriminating ability or sub-optimal calibration of the prediction models, or the absence of external model validation. We therefore aimed to develop a prognostic prediction model of PJI within 90 days after THA that is based on easily accessible data that are available in the setting of clinical decision-making. Furthermore, we aimed to validate the model both internally in Sweden, as well as externally on a Danish national cohort.

@@

The derivation cohort was obtained from SHAR `r period`, with the choice of 2008 as the starting year because registration of ASA class and body mass index (BMI) was initiated then. Deterministic data linkage was achieved by using personal identity numbers that are assigned to all Swedish residents at birth or immigration[@Ludvigsson2009; @Cnudde2016]. The diagnosis underlying the index THA procedure (categorized into `r glue::glue_collapse(tolower(levels(df$P_DiaGrp)), ", ", last = " and ") %>% {gsub("\\(avn\\)", "[AVN]", .)}`), type of THA fixation (cemented or uncemented stem and cup respectively), age at index surgery, sex, BMI prior to surgery, ASA class at the time of surgery, type of hospital, and year of surgery, were obtained from the SHAR. Data on comorbidities recognized during one year prior to index surgery, were obtained from the National Patient Register (NPR) where diagnoses based on in- and outpatient episodes in all private and public hospitals are coded according to ICD-10 together with admission and discharge dates. Data on marital status and the educational level of patients were obtained from the longitudinal integration database for health insurance and labour market studies (LISA) from Statistics Sweden. The SHAR has completeness of 96-98 % and 100 % coverage (82 hospitals in 2020)[@Karrholm2018], the LISA has coverage of 98 % with 85 % accuracy, and completeness for the NPR is above 99 % and the positive predictive value is 85-95 %. Death dates were linked from the national population register. Only the last operated hip was considered for patients with bilateral THA[@Bulow2020]. Patients with missing information on ASA class, or an ASA class of IV and above, were excluded, as were patients with missing information on BMI or a measurement above 50. Patients with missing information on educational levels or the type of hospital performing index surgery were also excluded, as were children under the age of 18.

@@

All surgery related variables, ASA and BMI were obtained from the DHR. The DHR is a clinical quality database established in 1995 on all primary and revision THA surgeries, with registration completeness of more than 95 % during the last decade for both primary and revision THAs[@Gundtoft2016]. All hospitals performing THA in Denmark (40 in 2019) are included in the register. 
The Danish National Patient Register holds data on ICD-10 codes as well as all procedures codes for hospitalizations and outpatient visits at Danish public and private hospitals[@Schmidt2015]. Data on education and civil status were obtained from Statistics Denmark[@zotero-2587].

@@

The Charlson and Elixhauser comorbidity indices are based on the presence of ICD-10 codes grouped into `r nrow(coder::charlson)` and `r nrow(coder::elixhauser)` categories each[@Quan2005]. Such codes were identified from reported in- and outpatient hospital visits recorded in the national patient registers during one year prior to surgery. Diagnostic categories from those indices were combined to define `r nrow(tab_categorization)` categories of comorbidity that were chosen in order to be easily identifiable in a clinical setting (Supplementary Tab. \@ref(tab:tabcategorization)).

@@

PJI was defined by the occurrence of relevant ICD-10 or procedural Nordic Medico-Procedural Committee (NOMESCO)-codes recorded in the national patient registers for hospital visits within 90 days of THA (Supplementary Tab. \@ref(tab:infcodes)), or if the patient was reported to SHAR or DHR as re-operated due to infection within the same time frame.

@@

Model development was performed by bootstrap ranking and a logistic least absolute shrinkage and selection operator (LASSO)[@Guo2015; @Baranowski2020; @Garland2021]. Age was normalized prior to modelling, and the estimated coefficients thus indicated variable importance on the same scale as categorical variables. BMI was divided into four categories based on the six intervals proposed by the World Health Organization (WHO; BMI $<25$: under- or normal weight, $25-29.9$: overweight, $30-34.9$: class I obesity, and BMI $\geq 35$: class II-III obesity).

Ten-fold cross validation was performed with a range of potential penalty values ($\lambda_i$:s) in the LASSO model, keeping the $\lambda$ that minimized the mean cross-validated deviance. 100 bootstrap samples were then drawn from the observed data set.

@@

Univariable prediction models based on either ASA class, the Charlson, Elixhauser, or the Rx Risk V comorbidity indices were fitted for comparison, as well as multivariable models combining those measures with age, sex and BMI (10 models in total). The Rx Risk V index is a comorbidity index based on medical prescriptions coded by the Anatomical Therapeutic Chemical (ATC) classification system during one year prior to surgery. This data was retrieved from the Swedish and Danish[@Kildemoes2011] national prescription registers and classified according to Pratt et al[@Pratt2018].

@@

The main, reduced and the 10 reference models were used to predict the probability of PJI within 90 days for patients derived from the SHAR (internal validation). Receiver operating characteristic (ROC) curves and their corresponding areas under the curve (AUC) were used to describe discrimination with 95 % confidence intervals. The bias-corrected Somers' $D_{xy}$ rank correlation based on 100 resamples was used to adjust for optimism, but the correction for optimism only affected the third decimals in the presented results, and will therefore not be further discussed. Predicted probabilities and observed proportions were plotted with 95 % confidence intervals to graphically assess model calibration[@Nattino2016]. Coefficients of determination were estimated by Nagelkerke's $R^2$ for each model. The baseline model with the lowest $R^2$ ($R^2_0$) was compared to the other $k = 1, \ldots, 11$ models, by $1 - R^2_0 / R^2_k$. This is the fraction of new information, or the proportion of added explainable variation for model $k$ compared to baseline[@Harrell2018].

@@

We used `r R.version.string` (R Foundation for Statistical Computing, Vienna, Austria) with significant packages coder[@Bulow2020b], decoder, furrr, pROC, rms, tidymodels and tidyverse. We build a shiny-based web calculator to aid clinical usage (https://erikbulow.shinyapps.io/thamortpred/).

@@

Results from the 100 bootstrap replicates for variable selection were mostly coherent. Diagnosis, BMI and some comorbidities where selected each time. Age was selected 82 out of 100 times, but some factors, such as civil status and type of hospital, were only sporadically selected (Supplementary Tab. \@ref(tab:brlprop)).

@@

Both Sweden and Denmark have strived for mass higher education[@Isopahkala-Bouret2018]. This is, to a higher degree achieved via the university sector in Sweden, while the educational system is more diverse in Denmark.[@Kim2002] National comparisons of educational levels are therefore difficult, since the Danish student register does not include adult- or continuing education[@zotero-2598]. Patient distribution among ASA classes were similar in both countries, but the comorbidity burden seems to differ. ICD-10 and NOMESCO codes are recorded for both in- and outpatient hospital visits in the national patient registries in both countries, but the recording of (optional) secondary diagnoses is 23 % in Denmark[@Schmidt2015], compared to 47 % in the Swedish cohort. Serious comorbidities such as cancer, are therefore reported equally in both countries, while for example arterial hypertension, is less documented in Denmark. Also, patients with chronic conditions, such as diabetes, are more commonly treated in primary care settings in Denmark, and are therefore less reported to their national patient register[@Schmidt2015]. The RxRisk V index is a weighted index, which is difficult to compare without further analysis. Prescription patterns might differ by indications based on different national guidelines, different subsidization, and different medications sold over-the-counter. Also, the Danish, but not the Swedish, national prescription register, includes prescriptions from nursery homes and other long-term care facilities. Taken together, the similarities between the countries might be more pronounced than indicated by Tab. \@ref(tab:tab1).

However, the difference in the mode of THA fixation, with a dominance of cemented components in Sweden and much more widespread usage of uncemented components in Denmark is tangible and has been described elsewhere[@Bunyoz2020].

@@

Cancer seems to confer an increased risk of PJI in several observational studies[@Bozic2012; @Everhart2013; @Poultsides2013; @Wingert2016a], but this comorbidity was not included in our model. It nevertheless appeared in `r subset(coefs_selected_tab, variable == "Cancer", "n", drop = TRUE)` of the 100 bootstrap replicates (Tab. \@ref(tab:brlprop)), and could have been included in a less parsimonious model. Neither diabetes was included in our model, but has been in previous risk calculators[@Wingert2016a; @Tan2018; @Inacio2015]. On the other hand, diabetes is not consistently found to be associated with the risk of surgical site infection or PJI after THA surgery[@Namba2012].

Information on peri-operative hyperglycaemia, morning blood glucose, HBA1c[@Mraovic2011; @Iorio2012], or low preoperative haemoglobin,[@Rasouli2014a] could have improved our model. But although we had access to the ICD codes defining the presence of anaemia, we had no information on actual laboratory findings. Use of subcutaneously applied Tumour Necrosis Factor-alpha blockers[@Gilson2010; @Momohara2011], intra-articular steroid application[@Cancienne2015] or operating time[@Maoz2015; @Willis-Owen2010; @Berbari2012], are also associated with the risk of PJI after THA. However, we chose to only consider factors available before surgery and are thus able to use the model as a tool for shared decision making using the implemented web calculator. This is a strength of our model.

We also confirmed that, although mentioned as risk factors for surgical site infections after joint arthroplasty[@Rasouli2014a; @Inacio2015a], prediction models based on the comorbidity indices developed by Charlson or Elixhauser and the RxRisk V index, resulted in poor predictive power. This agrees with previous findings describing that these measures of comorbidity rather poorly predict mortality and patient reported outcomes after THA[@Bulow2017; @Bulow2019]. However, when used in conjunction with information on age and sex, the AUC for these comorbidity indices increased, although never quite reaching the AUC of our model (Fig. \@ref(fig:aucci) and Supplementary Tab. \@ref(tab:fni)).

@@

The selected variables are nevertheless coherent with previous studies. Patients with primary osteoarthritis seem less prone to developing PJI than patients receiving THA for other reasons[@Pedersen2010a; @Dale2012; @Bozic2012; @Inacio2015a]. Obesity is associated with an increased risk of reoperations or adverse events after THA[@Sayed-Noor2019] and with an increased risk of developing surgical site infections or PJI after total hip or knee arthroplasty[@Malinzak2009; @Jamsen2012; @Maoz2015; @Everhart2013; @Tan2018]. CNS disease (cerebrovascular disease, dementia, hemiplegia, and Parkinson's disease) has also been previously associated with PJI[@Bozic2012], as have male sex[@Dale2012; @Wingert2016a; @Tan2018], liver disease[@Deleuran2015] and ASA class[@Berbari2012; @Pulido2008; @Namba2012].

THA fixation using cement without antibiotics confers an increased risk of PJI[@Dale2012], but such cement brands were not in clinical use in Sweden during the studied period. We found no indication that the type of fixation was associated with the risk of PJI, and the mode of THA fixation was also not selected as a potential predictor variable in any of the iterations during model development.

@@

@@

# Author contributions

# Role of funding sources

No external funding

# Conflict of interest

@@

<!-- TABLES -->

@@

## Supplementary tables
